\documentclass[11pt]{article}

\usepackage{amsmath, amssymb, amsthm}
\usepackage{mathtools}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{xcolor}

\geometry{margin=1in}

\newtheorem{proposition}{Proposition}
\newtheorem{remark}{Remark}

\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\diag}{diag}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\dd}{\,\mathrm{d}}
\newcommand{\bx}{\mathbf{x}}
\newcommand{\bs}{\mathbf{s}}
\newcommand{\bphi}{\boldsymbol{\phi}}

\title{Bayesian Inference for Inhomogeneous Poisson Processes\\
via Adaptive FEM Triangulation and the Mat\'ern SPDE}
\author{Jack Heinzel}
\date{Draft --- \today}

\begin{document}
\maketitle

\begin{abstract}
We describe a computational pipeline for inferring the intensity function of an
inhomogeneous Poisson process over a $d$-dimensional domain.
The log-intensity is modelled as a Gaussian random field whose prior is derived
from the stochastic partial differential equation (SPDE) representation of the
Mat\'ern covariance family \citep{lindgren2011}.
The field is discretised on an adaptive triangulation driven by the observed
point pattern, assembled with finite element methods (FEM), and sampled jointly
with the hyperparameters $(\kappa, \sigma, \mu)$ using the No-U-Turn Sampler
(NUTS) implemented in BlackJAX.
We discuss several implementation subtleties: the domain-area correction for
buffer nodes, an $O(n)$ eigenvalue trick for the log-determinant, and a centred
parameterisation that eliminates Neal's funnel.
\end{abstract}

\tableofcontents
\bigskip

% ============================================================
\section{Problem Setting}
% ============================================================

Let $\mathcal{D} \subset \R^d$ be a bounded domain and let
$\{s_1, \ldots, s_n\} \subset \mathcal{D}$ be a realisation of an
\emph{inhomogeneous Poisson process} with intensity function $\lambda(s) \ge 0$.
The likelihood for the observed point pattern is
\begin{equation}
  \log p(\{s_i\} \mid \lambda) = \sum_{i=1}^n \log \lambda(s_i) - \int_{\mathcal{D}} \lambda(s) \dd s.
  \label{eq:poisson-ll}
\end{equation}
Our goal is Bayesian inference for $\lambda(\cdot)$ given only the $n$
observed locations.  We adopt the \emph{Log-Gaussian Cox Process} (LGCP)
model \citep{moller1998lgcp}, in which
\begin{equation}
  \log \lambda(s) = \mu + x(s), \qquad x \sim \mathcal{GP}(0, \sigma^2 C_\nu(\cdot,\cdot;\kappa)),
  \label{eq:lgcp}
\end{equation}
where $\mu \in \R$ is a global mean (intercept), $x$ is a zero-mean Gaussian
random field with Mat\'ern covariance $C_\nu$ parameterised by range $\kappa$
and marginal standard deviation $\sigma$.

% ============================================================
\section{The Mat\'ern SPDE Prior}
% ============================================================

\subsection{Continuous SPDE}

\citet{lindgren2011} show that a Gaussian random field with Mat\'ern covariance
of smoothness $\nu$ and range $1/\kappa$ is the unique stationary solution of the
SPDE
\begin{equation}
  (\kappa^2 - \Delta)^{\alpha/2} \, x(s) = \mathcal{W}(s), \quad s \in \R^d,
  \label{eq:spde}
\end{equation}
where $\mathcal{W}$ is Gaussian white noise, $\Delta$ is the Laplacian, and
$\alpha = \nu + d/2$.  For $d=2$ the integer case $\alpha=2$ ($\nu=1$) is most
common and is the focus of this work.

\subsection{Generalised Anisotropic SPDE}

The stiffness operator can be generalised to allow spatially varying diffusion
\begin{equation}
  (\kappa(s)^2 - \nabla \cdot H(s) \nabla)^{\alpha/2} \, x(s) = \mathcal{W}(s),
  \label{eq:spde-gen}
\end{equation}
where $H(s) \in \R^{d\times d}$ is a positive-definite diffusion tensor.
Setting $H = I$ and $\kappa$ constant recovers \eqref{eq:spde}.

\subsection{Marginal Variance}

For the stationary isotropic case in $d=2$ with $\alpha=2$, the marginal
variance of the field is
\begin{equation}
  \sigma_x^2 = \frac{1}{4\pi\kappa^2}.
  \label{eq:margvar}
\end{equation}
For $\kappa = 6$ this gives $\sigma_x \approx 0.047$, far too small for
intensity fields that vary by one or more log-units.  Including a free scale
parameter $\sigma$ (Section~\ref{sec:hyperparams}) is therefore essential.

% ============================================================
\section{Adaptive Triangulation}
% ============================================================

\subsection{Recursive Equal-Count kd-Tree}

We build a mesh that is automatically denser where the observed point pattern is
dense.  Starting from a root cell covering the domain, we recursively split
each cell along its longest axis at the data median until every leaf cell
contains at most $n_{\max}$ points (the refinement threshold).  This is a
\emph{balanced} kd-tree partition: each split divides the data approximately
evenly, so the resulting leaf sizes are $O(n/n_{\rm leaves})$.

The split value is clamped away from the cell boundary by a relative tolerance
$\epsilon = 10^{-8}$ of the cell width, ensuring no degenerate child cell.

\subsection{Mesh Nodes and Delaunay Triangulation}

Each leaf cell contributes one node to the mesh: the centroid of its data
points if the cell is non-empty, or the geometric cell midpoint otherwise.
The $2^d$ corners of the bounding box are added as fixed boundary nodes.

The Delaunay triangulation of all nodes defines the mesh $\mathcal{T}$.
It is rebuilt from scratch after each refinement or coarsening step.

\subsection{Buffer Region}

To ensure the triangulation is regular near the domain boundary, the bounding
box is taken slightly larger than the observation domain---typically
$[-0.1, 1.1]^2$ when observations lie in $[0,1]^2$.  This introduces
\emph{buffer nodes} in the extended region.  We account for them explicitly in
the likelihood (Section~\ref{sec:obs-bounds}).

% ============================================================
\section{FEM Discretisation}
% ============================================================

\subsection{Galerkin Weak Form}
\label{sec:galerkin}

The Galerkin method turns the SPDE into a finite-dimensional linear system
in two steps.

\textbf{Step 1 --- approximate the field.}  Represent $x(s)$ as a linear
combination of hat basis functions:
\begin{equation}
  x(s) \approx \sum_{j=1}^m x_j\,\phi_j(s),
  \label{eq:fem-approx}
\end{equation}
reducing the unknown field to $m$ nodal values $\mathbf{x}\in\R^m$.

\textbf{Step 2 --- enforce the PDE weakly.}  Rather than requiring the SPDE
to hold pointwise (which would be too strong for the piecewise-linear
approximation), we require the residual to be orthogonal to every basis
function $\phi_i$.  For the $\alpha=1$ SPDE
$(\kappa^2-\Delta)x=\mathcal{W}$, this means
\begin{equation}
  \int_\mathcal{D} \phi_i(s)\,\bigl[(\kappa^2-\Delta)x(s)\bigr]\dd s
  = \int_\mathcal{D} \phi_i(s)\,\mathcal{W}(s)\dd s, \quad i=1,\ldots,m.
  \label{eq:galerkin-condition}
\end{equation}
The Laplacian term is handled by \emph{integration by parts}
(Green's first identity):
\begin{equation}
  \int_\mathcal{D} \phi_i\,(-\Delta x)\dd s
  = \int_\mathcal{D} \nabla\phi_i\cdot\nabla x\dd s,
  \label{eq:ibp}
\end{equation}
where the boundary term vanishes.  This step is essential: $\phi_i$ is
piecewise linear and its second derivatives are zero almost everywhere, so the
left side of \eqref{eq:ibp} would be unusable; the right side requires only
first derivatives, which are constant and well-defined on each simplex.

Substituting \eqref{eq:fem-approx} into the left-hand side of
\eqref{eq:galerkin-condition} and collecting:
\begin{equation}
  \sum_j x_j \Bigl[\kappa^2\!\int\!\phi_i\phi_j\dd s
                  +\int\!\nabla\phi_i\cdot\nabla\phi_j\dd s\Bigr]
  = \sum_j x_j\,[\kappa^2 C_{ij} + G_{ij}]
  = [K\mathbf{x}]_i.
  \label{eq:lhs}
\end{equation}
For the right-hand side, projecting white noise onto $\phi_i$ gives
$z_i = \langle\phi_i,\mathcal{W}\rangle$ with
$\mathrm{Cov}(z_i,z_j)=\langle\phi_i,\phi_j\rangle=C_{ij}$,
so $\mathbf{z}\sim\mathcal{N}(0,C)$.

The Galerkin system is therefore the discrete stochastic equation
\begin{equation}
  K\mathbf{x} = \mathbf{z}, \qquad \mathbf{z}\sim\mathcal{N}(0,C),
  \label{eq:galerkin-system}
\end{equation}
with $K=\kappa^2 C+G$ assembling the two bilinear forms identified in
\eqref{eq:lhs}.  Sections~\ref{sec:mass}--\ref{sec:stiffness} define $C$ and $G$
explicitly; Section~\ref{sec:spde-op} derives the precision matrix from
\eqref{eq:galerkin-system}.

\subsection{Basis Functions and Local Geometry}

On the triangulation $\mathcal{T}$ with $m$ nodes and $T$ simplices, we use
piecewise linear \emph{hat} basis functions $\{\phi_i\}_{i=1}^m$.  The hat
function $\phi_j$ equals 1 at node $j$, 0 at every other node, and varies
linearly in between.  Within any single simplex it is a flat tilted plane, so
its gradient is \emph{constant} on each simplex and can be computed from
geometry alone.

\paragraph{Reference simplex and the map $B_\tau$.}
Every computation is carried out on a fixed \emph{reference simplex}
$\hat\tau$ with vertices at the origin $\hat{v}_0=\mathbf{0}$ and the
standard basis vectors $\hat{v}_k = e_k$ for $k=1,\ldots,d$.  On this
template the hat functions are trivially
\[
  \hat\phi_0(\hat s) = 1 - \hat s_1 - \cdots - \hat s_d, \qquad
  \hat\phi_k(\hat s) = \hat s_k \;\;(k\ge 1),
\]
with reference gradients $\nabla_{\hat s}\hat\phi_j = \hat{e}_j$, where
$\hat{e}_0 = -\mathbf{1}$ and $\hat{e}_k$ is the $k$-th standard basis
vector.

The matrix
\begin{equation}
  B_\tau = [v_1 - v_0, \;\ldots\;, v_d - v_0] \in \R^{d\times d}
\end{equation}
maps the reference simplex to the physical simplex $\tau$ via
$s = v_0 + B_\tau\hat s$.  Its columns are the edge vectors of $\tau$
emanating from $v_0$, so $B_\tau$ encodes the shape and orientation of the
triangle.  The volume follows immediately: $|\tau| = |\det B_\tau|/d!$.

\paragraph{Why $B_\tau^{-\top}$ for gradients.}
The gradient $\nabla_s\phi_j^{(\tau)}$ is the vector $g$ satisfying
$g\cdot\delta s = \delta\phi$ for every physical displacement $\delta s$.
Writing $\delta s = B_\tau\,\delta\hat s$ and using the reference gradient
$\nabla_{\hat s}\hat\phi_j = \hat{e}_j$:
\[
  g \cdot B_\tau\,\delta\hat s
  = \hat{e}_j \cdot \delta\hat s
  \quad\text{for all }\delta\hat s,
\]
which gives $B_\tau^{\top}g = \hat{e}_j$, and therefore
\begin{equation}
  \nabla_s \phi_j^{(\tau)} = B_\tau^{-\top}\hat{e}_j, \quad j = 0, \ldots, d.
  \label{eq:grad-phi}
\end{equation}
Intuitively, $B_\tau^{-1}$ converts a physical direction into reference
coordinates, and the transpose accounts for the fact that gradients scale
inversely with distances (a squeezed triangle has steeper slopes).  Because
$B_\tau$ is constant on $\tau$, so is $\nabla_s\phi_j^{(\tau)}$: the hat
function is a flat ramp, and \eqref{eq:grad-phi} gives its slope.

\subsection{Mass Matrix}
\label{sec:mass}

The mass matrix $C \in \R^{m\times m}$ has entries
\begin{equation}
  C_{ij} = \int_\mathcal{D} \phi_i(s)\phi_j(s) \dd s
         = \sum_{\tau \ni i,j} |\tau| \frac{1 + \delta_{ij}}{(d+1)(d+2)}.
  \label{eq:mass}
\end{equation}
$C_{ij}$ measures the \emph{overlap} between hat functions $\phi_i$ and
$\phi_j$: it is nonzero only when nodes $i$ and $j$ share at least one
simplex.  The diagonal entry $C_{ii}$ is proportional to the total area of
all simplices containing node $i$.  The matrix is symmetric positive definite
with sparsity matching the mesh topology.

The \emph{row sum} $\tilde{c}_j = (C\mathbf{1})_j = \sum_k C_{jk}$ is the
total area ``owned'' by node $j$: it equals $\sum_{\tau\ni j}|\tau|/(d+1)$,
the average share of each incident simplex.  Geometrically, $\tilde{c}_j$
approximates the area of the Voronoi cell\footnote{It's not really a Voronoi cell, but it's close-ish to one depending on the triangulation.} around node $j$.  The diagonal
matrix $\tilde{C} = \diag(C\mathbf{1})$ is the \emph{lumped} mass matrix,
used in Section~\ref{sec:integral-approx} to approximate the Poisson integral
as a weighted sum $\sum_j \tilde{c}_j \lambda_j$.

In the SPDE operator $(\kappa^2 - \Delta)$, the term $\kappa^2$ acts like a
scalar multiple of the identity on functions.  After Galerkin projection onto
the hat-function basis, this becomes $\kappa^2 C$: the mass matrix is the
FEM discretisation of the identity operator, weighted by local area.

\subsection{Stiffness Matrix}
\label{sec:stiffness}

The stiffness matrix $G \in \R^{m\times m}$ has entries
\begin{equation}
  G_{ij} = \int_\mathcal{D} \nabla\phi_i(s) \cdot \nabla\phi_j(s) \dd s
         = \sum_{\tau \ni i,j} |\tau| \, \nabla\phi_i^{(\tau)} \cdot \nabla\phi_j^{(\tau)}.
  \label{eq:stiff}
\end{equation}
$G_{ij}$ measures how much nodes $i$ and $j$ \emph{resist spatial variation}
together: it is the inner product of the slopes of the two hat functions.
By integration by parts, $\mathbf{x}^T G\, \mathbf{x} =
\int|\nabla x(s)|^2\dd s$, so the quadratic form in $G$ is exactly the total
squared-gradient energy of the piecewise-linear field
$x(s)=\sum_j x_j\phi_j(s)$.  Smooth fields (small gradients everywhere) have
low $G$-energy; rough or wiggly fields have high $G$-energy.

Equivalently, $G$ discretises the negative Laplacian $-\Delta$: it can be
thought of as a network of \emph{springs} connecting adjacent nodes, where
$G_{ij}$ is (minus) the spring constant for the edge between $i$ and $j$.
For a Delaunay mesh in 2D, $G_{ij} \le 0$ for all neighbours
($i\neq j$), and the action $(G\mathbf{x})_i = -\sum_j G_{ij}(x_j - x_i)$
sums the spring forces pulling node $i$ toward its neighbours' values.

Together, $K = \kappa^2 C + G$ discretises $(\kappa^2 - \Delta)$: the mass
term $\kappa^2 C$ penalises large \emph{absolute} values of the field, and
the stiffness term $G$ penalises spatial \emph{roughness}.  Large $\kappa$
makes the mass term dominate, producing a field with short correlation length
(rapidly decorrelated); small $\kappa$ lets the stiffness term dominate,
producing a smoother field correlated over longer distances.

For the generalised case \eqref{eq:spde-gen}, the integrand becomes
$\nabla\phi_i^T H(\bar\tau) \nabla\phi_j$ where $H$ is evaluated at the
simplex centroid $\bar\tau$: the anisotropic diffusion tensor $H$ stretches
the spring network directionally, making correlations longer in some
directions than others.

\subsection{SPDE Operator and Precision Matrix}
\label{sec:spde-op}

The matrix $K = \kappa^2 C + G$ is the FEM discretisation of the operator
$(\kappa^2 - \Delta)$, as assembled in \eqref{eq:lhs}.

\paragraph{Precision for $\alpha=1$.}
From the Galerkin system \eqref{eq:galerkin-system}, $\mathbf{x}=K^{-1}\mathbf{z}$
with $\mathbf{z}\sim\mathcal{N}(0,C)$, so the covariance of $\mathbf{x}$ is
$K^{-1}CK^{-1}$ ($K$ is symmetric) and the precision is
\begin{equation}
  Q_{\alpha=1} = K C^{-1} K.
  \label{eq:Q1}
\end{equation}

\paragraph{Precision for $\alpha=2$.}
The $\alpha=2$ SPDE $(\kappa^2-\Delta)^2 x = \mathcal{W}$ factors as two
applications of $(\kappa^2-\Delta)$.  Applying the Galerkin discretisation
twice gives the same form $Q = KC^{-1}K$, but $C^{-1}$ is dense (the inverse
of a sparse matrix), making it expensive to work with.  Following
\citet{lindgren2011}, we replace $C$ by the \emph{lumped} mass matrix
$\tilde{C}=\diag(C\mathbf{1})$:
\begin{equation}
  Q = K \, \tilde{C}^{-1} \, K.
  \label{eq:Q}
\end{equation}
Because $\tilde{C}$ is diagonal, $\tilde{C}^{-1}$ costs nothing to apply.
The substitution is a consistent approximation: $\tilde{C}\to C$ as the mesh
refines (the lumped and consistent mass matrices converge).  The resulting $Q$
is sparse --- its nonzero pattern is that of $K^2$ restricted to the mesh
graph --- and positive definite for any $\kappa>0$, making the prior proper.

For general $\alpha\ge 1$,
\begin{equation}
  Q = K\bigl(\tilde{C}^{-1}K\bigr)^{\alpha-1}.
  \label{eq:Q-general}
\end{equation}

% ============================================================
\section{Likelihood Discretisation}
% ============================================================

\subsection{Observation Matrix}

For observation $s_i$ lying in simplex $\tau(i)$ with vertices
$v_0^{(i)}, \ldots, v_d^{(i)}$, let $\lambda_i = (\lambda_{v_0^{(i)}}, \ldots,
\lambda_{v_d^{(i)}})$ be the nodal values.  By piecewise linear interpolation,
\begin{equation}
  \log\hat\lambda(s_i) = \mu + \sum_{j=0}^{d} b_{ij} \, x_{v_j^{(i)}},
\end{equation}
where $b_{ij} \ge 0$ are the barycentric coordinates of $s_i$ in $\tau(i)$.

The sum term in \eqref{eq:poisson-ll} becomes $\sum_i x_{v_j^{(i)}} b_{ij} = A\bx$
where $A \in \R^{n \times m}$ is the sparse barycentric projection matrix.
Because the $i$-th row of $A$ is the barycentric coordinates of $s_i$, and
since barycentric coordinates sum to one, the log-likelihood can be written as
\begin{equation}
  \sum_{i=1}^n \log\lambda(s_i) = n\mu + \mathbf{d}^T \bx,
  \label{eq:data-term}
\end{equation}
where the \emph{data-accumulation vector} $\mathbf{d} \in \R^m$ is
$d_j = \sum_{i : v_j \in \tau(i)} b_{ij}$, assembled by scattering barycentric
weights to their respective nodes.

\subsection{Integral Approximation and Lumped Mass Quadrature}

The integral $\int_\mathcal{D} \lambda(s) \dd s$ is approximated by a
nodal quadrature rule.  Expanding $\lambda(s) = \exp(\mu + x(s))$ and
integrating the piecewise linear interpolant gives
\begin{equation}
  \int_\mathcal{D} \exp\!\bigl(\mu + x(s)\bigr) \dd s
  \approx \sum_{j=1}^m c_j \exp(\mu + x_j),
  \qquad c_j = \int_\mathcal{D} \phi_j(s) \dd s = \tilde{C}_{jj}.
  \label{eq:integral-approx}
\end{equation}
The lumped weights $c_j = \tilde{C}_{jj} = (C\mathbf{1})_j$ equal the area
contributed to node $j$ across all incident simplices, specifically
$c_j = \sum_{\tau \ni j} |\tau|/(d+1)$.

\subsection{Domain Correction for Buffer Nodes}
\label{sec:obs-bounds}

When the mesh bounding box $\hat{\mathcal{D}}$ is larger than the observation
domain $\mathcal{D}_{\rm obs}$, the buffer nodes $j \notin \mathcal{D}_{\rm obs}$
have $c_j > 0$ but lie outside the region over which the Poisson likelihood is
defined.  Including them inflates the integral, biasing $\mu$ downward.

The fix is to zero the lumped weights for buffer nodes before forming the integral:
\begin{equation}
  \tilde{c}_j = c_j \cdot \mathbf{1}[s_j \in \mathcal{D}_{\rm obs}].
  \label{eq:obs-bounds}
\end{equation}
With this correction, the likelihood integral is $\sum_j \tilde{c}_j \exp(\mu+x_j)$,
which integrates only over the observation domain.

\subsection{Posterior Normalization Property}
\label{sec:gamma-posterior}

\begin{proposition}
Under a flat prior $p(\mu) \propto 1$ and the LGCP likelihood \eqref{eq:poisson-ll}
with the quadrature \eqref{eq:integral-approx}, the conditional posterior of the
expected total count $N = \sum_j \tilde{c}_j \exp(\mu + x_j)$, given $\bx$, is
$\mathrm{Gamma}(n, 1)$ with mean $n$.
\end{proposition}

\begin{proof}
Let $Z(\bx) = \sum_j \tilde{c}_j e^{x_j}$.  The log-posterior in $\mu$,
$p(\mu \mid \bx, \{s_i\}) \propto \exp(n\mu - Z(\bx) e^\mu)$,
is the kernel of a $\mathrm{Gamma}(n, 1/Z(\bx))$ distribution on
$N = Z(\bx)e^\mu$.  Hence $\E[N \mid \bx] = n$ for any realization of
$\bx$.
\end{proof}

This property provides a useful diagnostic: after sampling, the posterior mean
of $\sum_j \tilde{c}_j \exp(\hat\mu + \hat x_j)$ should equal $n$.  Without
the domain correction~\eqref{eq:obs-bounds}, the guarantee holds for the full
mesh domain $\hat{\mathcal{D}}$ rather than $\mathcal{D}_{\rm obs}$.

% ============================================================
\section{Joint Inference for Hyperparameters}
\label{sec:hyperparams}
% ============================================================

\subsection{Parameterisation and Prior}

We infer $(\mu, \kappa, \sigma, \bx)$ jointly.  The model is
\begin{align}
  \log\kappa &\sim \mathcal{N}(\mu_\kappa, \sigma_\kappa^2), \\
  \log\sigma  &\sim \mathcal{N}(\mu_\sigma, \sigma_\sigma^2), \\
  \bx \mid \kappa, \sigma &\sim \mathcal{N}(0,\; \sigma^2 Q(\kappa)^{-1}), \\
  \{s_i\} \mid \mu, \bx &\sim \text{LGCP with } \log\lambda = \mu + \bx.
  \label{eq:model}
\end{align}

\subsection{Centred Parameterisation to Avoid Neal's Funnel}
\label{sec:centred}

A non-centred parameterisation $\bx = \sigma \tilde\bx$ with
$\tilde\bx \sim \mathcal{N}(0, Q^{-1})$ creates a \emph{Neal's funnel}
\citep{neal2003slice}: the joint distribution
$({\sigma, \tilde\bx})$ and $(2\sigma, \tilde\bx/2)$ produce nearly identical
likelihood values, yet the prior on $\tilde\bx$ \emph{improves} as
$\tilde\bx \to 0$, giving the sampler a free gradient towards $\sigma \to \infty$
with $\tilde\bx \to 0$.  Traces of $\log\sigma$ diverge.

The \emph{centred parameterisation} resolves this by working directly with
$\bx$ in \eqref{eq:model}.  The prior log-density is
\begin{equation}
  \log p(\bx \mid \kappa, \sigma) =
    -\frac{1}{2\sigma^2} \bx^T Q \bx
    + \tfrac{1}{2}\log|Q|
    - m \log\sigma
    + \text{const},
  \label{eq:log-prior-x}
\end{equation}
where the $-m\log\sigma$ term (the Jacobian for the change-of-variables
$\bx \to \tilde\bx = \bx/\sigma$) strongly penalises large $\sigma$.
For $m \approx 130$ nodes, increasing $\sigma$ by a factor of 10 costs
$130 \log 10 \approx 299$ log-units in prior density.

\subsection{Log-Determinant via Generalised Eigenvalues}
\label{sec:logdet}

The log-prior \eqref{eq:log-prior-x} requires $\log|Q(\kappa)| = 2\log|K(\kappa)| - \log|\tilde{C}|$.
Recomputing $\log|K|$ at every MCMC step via Cholesky costs $O(m^3)$ per
iteration.  We reduce this to $O(m)$ via the spectral decomposition.

\begin{proposition}[Eigenvalue log-det trick]
Let $G v_i = \mu_i C v_i$ be the generalised eigenproblem (computed once,
$O(m^3)$).  Then for any scalar $\kappa$,
\begin{equation}
  \log|K(\kappa)| = \log|C| + \sum_{i=1}^m \log(\kappa^2 + \mu_i).
  \label{eq:logdet-trick}
\end{equation}
\end{proposition}

\begin{proof}
Factorise $K = \kappa^2 C + G = C(\kappa^2 I + C^{-1}G)$.  Then
$\log|K| = \log|C| + \log|\kappa^2 I + C^{-1}G|$.
The matrix $C^{-1}G$ has eigenvalues $\mu_i$, so
$\log|\kappa^2 I + C^{-1}G| = \sum_i \log(\kappa^2 + \mu_i)$.
\end{proof}

The eigenpairs $\{(\mu_i, v_i)\}$ are computed once before sampling via
\texttt{scipy.linalg.eigh}; within BlackJAX, \eqref{eq:logdet-trick} reduces
the log-determinant evaluation to a single sum over eigenvalues.

\begin{remark}
This trick requires scalar $\kappa$.  For non-stationary fields where
$\kappa(s)$ varies, the pencil $K(\kappa) = \diag(\kappa^2)C + G$ is no longer
a polynomial in a single scalar, and stochastic Lanczos quadrature (SLQ) would
be needed instead.
\end{remark}

\subsection{Full Log-Posterior}

Combining all terms, the log-posterior to be sampled is
\begin{align}
  \log p(\mu, \log\kappa, \log\sigma, \bx \mid \{s_i\})
  &= \underbrace{n\mu + \mathbf{d}^T \bx - \sum_j \tilde{c}_j e^{\mu+x_j}}_{\text{Poisson likelihood}} \notag \\
  &\quad - \underbrace{\frac{1}{2\sigma^2}\bx^T Q \bx + \frac{1}{2}\log|Q| - m\log\sigma}_{\text{SPDE prior on } \bx}
  \label{eq:log-posterior} \\
  &\quad - \underbrace{\frac{(\log\kappa - \mu_\kappa)^2}{2\sigma_\kappa^2}
         + \frac{(\log\sigma - \mu_\sigma)^2}{2\sigma_\sigma^2}}_{\text{hyperparameter priors}}. \notag
\end{align}

% ============================================================
\section{Stationary Anisotropic Extension}
\label{sec:anisotropic}
% ============================================================

\subsection{Anisotropy Matrix Parameterisation}

The isotropic model (Section~\ref{sec:hyperparams}) treats all directions
equally through a scalar range parameter $\kappa$.  A stationary
\emph{anisotropic} generalisation retains a constant diffusion tensor $H$ in
\eqref{eq:spde-gen}.  In $\R^2$ we parameterise
\begin{equation}
  H = R(\theta)\,\diag(\kappa_x^2,\,\kappa_y^2)\,R(\theta)^T,
  \label{eq:H-param}
\end{equation}
where $R(\theta)$ is the $2\times 2$ rotation matrix and $\kappa_x, \kappa_y > 0$
are effective range parameters along the principal axes of the field.
The elements of $H$ are
\begin{align}
  h_{11} &= \kappa_x^2\cos^2\!\theta + \kappa_y^2\sin^2\!\theta, \nonumber\\
  h_{12} &= (\kappa_x^2 - \kappa_y^2)\cos\theta\sin\theta,           \label{eq:H-components} \\
  h_{22} &= \kappa_x^2\sin^2\!\theta + \kappa_y^2\cos^2\!\theta.    \nonumber
\end{align}
The three parameters $(\log\kappa_x,\, \log\kappa_y,\, \theta)$ replace the
single scalar $\log\kappa$ from the isotropic model; the marginal variance
$\sigma$ remains a scalar.  The isotropic case is recovered by setting
$\kappa_x = \kappa_y$ (any $\theta$), yielding $H = \kappa_x^2 I$.

\subsection{FEM Basis Decomposition}
\label{sec:aniso-fem}

The anisotropic stiffness matrix
\[
G_{H,ij} = \int_\mathcal{D} \nabla\phi_i^T H\,\nabla\phi_j\,\dd s
\]
decomposes linearly over three basis matrices:
\begin{equation}
  G_H = h_{11}\,G_{xx} + h_{12}\,G_{xy} + h_{22}\,G_{yy},
  \label{eq:GH-decomp}
\end{equation}
with
\begin{align}
  G_{xx,ij} &= \sum_\tau |\tau|\,
    \partial_x\phi_i^{(\tau)}\,\partial_x\phi_j^{(\tau)}, \\
  G_{yy,ij} &= \sum_\tau |\tau|\,
    \partial_y\phi_i^{(\tau)}\,\partial_y\phi_j^{(\tau)}, \\
  G_{xy,ij} &= \sum_\tau |\tau|\!\left(
    \partial_x\phi_i^{(\tau)}\,\partial_y\phi_j^{(\tau)}
    + \partial_y\phi_i^{(\tau)}\,\partial_x\phi_j^{(\tau)}\right).
\end{align}
All three matrices are assembled once from the fixed mesh prior to sampling and
stored in COO sparse format.  The isotropic $G = G_{xx} + G_{yy}$ is the
special case $H = I$.

At each NUTS step, forming $G_H$ costs three sparse scalar-matrix additions, and
the SPDE operator
\begin{equation}
  K(H) = C + G_H(H)
  \label{eq:K-aniso}
\end{equation}
has the same sparsity pattern as the isotropic $K(\kappa)$.  Matrix--vector
products $K(H)\bx$ therefore cost $O(\mathrm{nnz}) \approx O(m)$ via COO
scatter-add, unchanged from the isotropic case.

\subsection{Log-Determinant Challenge}
\label{sec:logdet-aniso}

Section~\ref{sec:logdet} showed that for the isotropic operator $K(\kappa) =
\kappa^2 C + G$, the log-determinant reduces to $O(m)$ work per step once the
generalised eigenpairs $G v_i = \mu_i C v_i$ have been precomputed:
\[
  \log|K(\kappa)| = \log|C| + \textstyle\sum_i \log(\kappa^2 + \mu_i).
\]
This is possible because $K$ is a \emph{scalar-affine} function of the single
parameter $\kappa^2$: the eigenvectors $\{v_i\}$ are fixed and only the
eigenvalues shift.

For the anisotropic operator \eqref{eq:K-aniso}, the matrix
$K(H) = C + h_{11}G_{xx} + h_{12}G_{xy} + h_{22}G_{yy}$
is a linear combination of four sparse matrices whose coefficients
$(h_{11}, h_{12}, h_{22})$ all change at every NUTS step.  No fixed eigenbasis
can diagonalise $K(H)$ for all $H$ simultaneously: a generalised eigenproblem
$K(H) v = \mu C v$ would need to be resolved whenever $(h_{11}, h_{12}, h_{22})$
changes.

Consequently, evaluating $\log|K(H)|$ requires a dense factorisation at each
leapfrog step,
\begin{equation}
  \log|K(H)| = \texttt{slogdet}\bigl(K(H)\bigr),
  \quad\text{cost } O(m^3) \text{ per step.}
  \label{eq:slogdet-aniso}
\end{equation}
The gradient required by NUTS,
\begin{equation}
  \frac{\partial\log|K|}{\partial\theta_k}
  = \tr\!\left(K^{-1}\frac{\partial K}{\partial\theta_k}\right),
  \label{eq:grad-logdet-aniso}
\end{equation}
requires an implicit $K^{-1}$ solve for each of the three anisotropy
parameters $(\log\kappa_x, \log\kappa_y, \theta)$, which JAX's automatic
differentiation of \texttt{jnp.linalg.slogdet} provides via the implicit
function theorem, again at $O(m^3)$ cost.

For meshes of size $m \lesssim 500$ the per-step $O(m^3)$ cost is acceptable
($\approx\!1$\,ms on a modern CPU), but it precludes scaling to
$m \gtrsim 2000$.  Finding an efficient $O(m)$--$O(m\log m)$ approximation
to $\log|K(H)|$ for parametric matrix pencils is an \textbf{open problem}.

\begin{remark}[Stochastic Lanczos quadrature]
One promising direction is \emph{stochastic Lanczos quadrature} (SLQ):
estimate $\log|K|$ as $m\cdot\E_{\mathbf{u}}[\log\tilde\mu(\mathbf{u})]$,
where $\tilde\mu$ is a stochastic Gauss quadrature approximation to the
spectral measure of $K$, requiring only matrix--vector products with $K$
(which remain $O(m)$).  The gradient \eqref{eq:grad-logdet-aniso} can be
obtained by differentiating the Gauss nodes and weights with respect to $H$.
This approach is left for future work.
\end{remark}

% ============================================================
\section{MCMC Sampling with NUTS}
% ============================================================

\subsection{BlackJAX and Window Adaptation}

We use the No-U-Turn Sampler \citep{hoffman2014nuts} as implemented in
BlackJAX \citep{blackjax2024}, a JAX-native library for Bayesian computation.
JAX's \texttt{jit}-compilation and automatic differentiation compute the
gradient of \eqref{eq:log-posterior} efficiently.

The step size and a diagonal mass matrix are tuned during a warm-up phase using
dual-averaging window adaptation.  After warm-up, samples are drawn and thinned
via a nested \texttt{jax.lax.scan} for memory efficiency.

\subsection{Initialisation}

The sampler is initialised at
\begin{equation}
  \mu^{(0)} = \log\!\left(\frac{n}{\sum_j \tilde{c}_j}\right),
  \quad \bx^{(0)} = \mathbf{0},
  \quad \log\kappa^{(0)} = \mu_\kappa,
  \quad \log\sigma^{(0)} = \mu_\sigma.
  \label{eq:init}
\end{equation}
The initialisation $\mu^{(0)}$ ensures the expected total count under the
prior equals $n$, consistent with Proposition~\ref{sec:gamma-posterior}.

\subsection{Posterior Prediction}

Given posterior samples $\{(\mu^{(s)}, \bx^{(s)})\}_{s=1}^S$, the intensity
at any evaluation point $s^* \in \mathcal{D}$ is obtained by barycentric
interpolation:
\begin{equation}
  \hat\lambda^{(s)}(s^*) = \exp\!\left(\mu^{(s)} + \sum_{j=0}^d b_j(s^*) x^{(s)}_{v_j(s^*)}\right),
\end{equation}
where $(v_0, \ldots, v_d)$ are the vertices of the containing simplex and
$(b_0, \ldots, b_d)$ are the barycentric coordinates of $s^*$.
Points outside the mesh return \texttt{NaN}.

% ============================================================
\section{Adaptive Triangulation via Reversible-Jump MCMC}
\label{sec:rjmcmc}
% ============================================================

\subsection{Refinement Tree}

We augment the static triangulation with a \emph{ternary refinement tree}.
Every triangle in the initial mesh is a root.  A \emph{birth move} splits a
leaf triangle $\tau^*$ at its barycenter $s^*$, replacing $\tau^*$ by the
three child triangles $(s^*,v_0,v_1)$, $(s^*,v_1,v_2)$, $(s^*,v_0,v_2)$.
A \emph{death move} is the exact reverse: if all three children of an
interior node $s^*$ are currently leaves, they are merged back into their
parent $\tau^*$ and $s^*$ is removed.  Birth and death are exact inverses,
so detailed balance is achievable.

Let $N_m$ be the number of leaf (birth-eligible) triangles and $M_m$ the
number of removable (death-eligible) nodes in mesh $m$.  After a birth on
$\tau^*$:
\begin{equation}
  N_{m'} = N_m + 2, \qquad
  M_{m'} = M_m + 1
           - \mathbf{1}\bigl[\text{all-leaf children test fails for parent}(s^*)\bigr],
  \label{eq:NM-update}
\end{equation}
where parent$(s^*)$ is the node whose earlier birth created $\tau^*$
(undefined for root triangles, so the indicator is 0 there).

\subsection{Conditional Prior Proposal}
\label{sec:rj-proposal}

After the 3-split the extended mesh $m'$ has $n_m+1$ nodes.  The SPDE prior
on the augmented field $(\mathbf{x}_m,x^*)$ is
$\mathcal{N}(0,\sigma^2 Q'^{-1})$ with $Q'=K'(\tilde{C}')^{-1}K'$.
Partitioning $Q'$ into the old nodes $(-*)$ and the new node $(*)$:
\begin{equation}
  Q' = \begin{pmatrix} Q'_{-*} & (Q'_{*-})^T \\ Q'_{*-} & Q'_{**} \end{pmatrix},
\end{equation}
the conditional prior of $x^*$ given the existing field is the
one-dimensional Gaussian
\begin{equation}
  x^*\mid\mathbf{x}_m,m',\kappa,\sigma
  \;\sim\; \mathcal{N}\!\left(\bar{x}^*,\;\frac{\sigma^2}{Q'_{**}}\right),
  \qquad
  \bar{x}^* = -\frac{Q'_{*-}\,\mathbf{x}_m}{Q'_{**}}.
  \label{eq:cond-prior-rj}
\end{equation}
Since $s^*$ connects only to $\{v_0,v_1,v_2\}$ in the mesh graph,
$K'_{*j}=0$ for $j\notin\{*,v_0,v_1,v_2\}$, and therefore
$Q'_{*j} = \sum_{k\in\{*,v_0,v_1,v_2\}} K'_{*k}(\tilde{c}'_k)^{-1}K'_{kj}$.
Both $Q'_{**}$ and $Q'_{*-}$ depend on $O(1)$ FEM values assembled from the
three new triangles; their evaluation is an $O(1)$ local computation.

\subsection{RJMCMC Acceptance Ratio}
\label{sec:rj-acceptance}

We propose $x^*$ from the conditional prior $q_*(x^*)=p(x^*\mid\mathbf{x}_m,m',\kappa,\sigma)$.
The standard \citet{green1995} acceptance ratio for the birth move is
\begin{equation}
  \alpha_\mathrm{b} = \min\!\left(1,\;
    \frac{\pi(m',\,\mathbf{x}_{m'})}
         {\pi(m,\,\mathbf{x}_m)}\cdot
    \frac{q_d\,N_m}{q_b\,M_{m'}\,q_*(x^*)}
  \right),
  \label{eq:rj-raw}
\end{equation}
where $q_b,q_d$ are the probabilities of attempting a birth or death move and
$\pi(m,\mathbf{x}_m)\propto p(m)\,p(\mathbf{x}_m\mid m,\kappa,\sigma)\,L(m,\mathbf{x}_m,\mu)$
is the unnormalised posterior.  (The Jacobian of the $u\to x^*$ transformation
cancels exactly against the proposal density by the change-of-variables
identity $q_*(x^*)\,|dx^*/du|^{-1}=\phi(u)$.)

\paragraph{Simplification via conditional prior proposal.}
Substituting $q_*(x^*)=p(\mathbf{x}_m,x^*\mid m')/p(\mathbf{x}_m\mid m')$
and cancelling common factors:
\begin{equation}
  \frac{p(\mathbf{x}_m,x^*\mid m')}
       {q_*(x^*)\,p(\mathbf{x}_m\mid m)}
  = \frac{p(\mathbf{x}_m\mid m')}{p(\mathbf{x}_m\mid m)}
  \;\eqqcolon\; \mathcal{R}_\mathrm{prior},
\end{equation}
so the acceptance ratio reduces to
\begin{equation}
  \alpha_\mathrm{b} = \min\!\left(1,\;
    \mathcal{R}_\mathrm{prior}
    \cdot
    \underbrace{\frac{L(m',\mathbf{x}_{m'},\mu)}
                     {L(m,\mathbf{x}_m,\mu)}}_{\mathcal{R}_L}
    \cdot
    \underbrace{\frac{p(m')}{p(m)}}_{\mathcal{R}_m}
    \cdot\frac{q_d\,N_m}{q_b\,M_{m'}}
  \right).
  \label{eq:rj-acc}
\end{equation}
The death move acceptance ratio is $\min(1,\alpha_\mathrm{b}^{-1})$.

\subsection{Prior Consistency and the Approximation $\mathcal{R}_\mathrm{prior}\approx 1$}
\label{sec:rj-prior}

The term $\mathcal{R}_\mathrm{prior} = p(\mathbf{x}_m\mid m')/p(\mathbf{x}_m\mid m)$
compares the SPDE prior on the \emph{old} nodes under the two meshes.
For a true Mat\'ern Gaussian process, this ratio equals 1 exactly: the GP
consistency property guarantees that the marginal at any fixed set of
locations is determined solely by the covariance function, independent of
which additional locations are included in the model.

For the \emph{FEM-discretised} prior, however, $\mathcal{R}_\mathrm{prior}\neq 1$
in general.  Splitting $\tau^*$ modifies the piecewise-linear basis functions
at $v_0,v_1,v_2$: the stiffness and mass integrals over the former $\tau^*$
are replaced by integrals over three smaller triangles, so $Q'_{-*}\neq Q$ and
the Schur complement $Q'_\mathrm{SC} \neq Q$.  The discrepancy is
\begin{equation}
  Q'_\mathrm{SC} - Q \;=\;
  \underbrace{(Q'_{-*} - Q)}_{\text{FEM change at }v_0,v_1,v_2}
  - \frac{(Q'_{*-})^T Q'_{*-}}{Q'_{**}},
  \label{eq:SC-diff}
\end{equation}
which has nonzeros only in rows/columns $\{v_0,v_1,v_2\}$ and their
2-hop neighbours.  Both terms scale with the area $|\tau^*|$ of the split
triangle, so $\|Q'_\mathrm{SC}-Q\|=O(|\tau^*|)$ and
$\log\mathcal{R}_\mathrm{prior}=O(|\tau^*|^2)$.  Finer triangles produce
smaller corrections, and the correction vanishes in the mesh-refinement limit
as the FEM prior converges to the true Matérn GP.

We therefore adopt the approximation
\begin{equation}
  \mathcal{R}_\mathrm{prior} \approx 1,
  \label{eq:Rprior-approx}
\end{equation}
which is exact for the true GP prior and is a consistent approximation for
the FEM prior.  The acceptance ratio simplifies to
\begin{equation}
  \boxed{
  \alpha_\mathrm{b} = \min\!\left(1,\;
    \underbrace{\frac{L(m',\mathbf{x}_{m'},\mu)}
                     {L(m,\mathbf{x}_m,\mu)}}_{\mathcal{R}_L}
    \cdot
    \underbrace{\frac{p(m')}{p(m)}}_{\mathcal{R}_m}
    \cdot\frac{q_d\,N_m}{q_b\,M_{m'}}
  \right).}
  \label{eq:rj-acc-simple}
\end{equation}

\begin{remark}[Exact cancellation via GP prior]
$\mathcal{R}_\mathrm{prior}=1$ holds \emph{exactly} if the prior on every mesh
is defined by the Mat\'ern covariance function directly---i.e.\
$p(\mathbf{x}_m\mid m,\kappa,\sigma)=\mathcal{N}(0,\Sigma_m)$ with
$\Sigma_{ij}=\sigma^2 C_\nu(s_i,s_j;\kappa)$---rather than by the FEM
precision.  In that case the proposal \eqref{eq:cond-prior-rj} becomes the
kriging predictor under the Mat\'ern covariance, replacing the local FEM
formula with a dense covariance solve.  The FEM approximation is then used
only for the likelihood (lumped-mass quadrature and barycentric interpolation),
not for the prior.
\end{remark}

\subsection{Likelihood Ratio}
\label{sec:rj-ll}

Because only the triangles touching $\tau^*$ change, the log-likelihood ratio
is local:
\begin{equation}
  \log\mathcal{R}_L
  = \!\sum_{i:\,s_i\in\tau^*}\!\!\!\log\frac{\hat\lambda'(s_i)}{\hat\lambda(s_i)}
  - \left[\tilde{c}'_{s^*}e^{\mu+x^*}
    + \sum_{j\in\{v_0,v_1,v_2\}}\!(\tilde{c}'_{v_j}-\tilde{c}_{v_j})\,e^{\mu+x_{v_j}}\right].
  \label{eq:rj-ll}
\end{equation}
The first sum covers only the $O(n/N_m)$ observations falling inside $\tau^*$,
using barycentric weights recomputed in the new child triangles.

\subsection{Mesh Prior}
\label{sec:mesh-prior}

A natural choice is a geometric prior on the number of leaf triangles:
\begin{equation}
  p(m) \propto \rho^{N_m-N_0}, \qquad 0<\rho<1,
  \label{eq:mesh-prior}
\end{equation}
where $N_0$ is the initial triangle count and $\rho$ controls the expected
depth of refinement.  Under \eqref{eq:mesh-prior}, $\mathcal{R}_m=\rho^2$
for every birth move (two new leaves added).

% ============================================================
\section{Infinite-Dimensional Sampling Methods}
\label{sec:inf-dim-samplers}
% ============================================================

The RJMCMC scheme of Section~\ref{sec:rjmcmc} handles the \emph{transdimensional}
part of the posterior — deciding how many mesh nodes to use.  Within a fixed mesh,
however, we still need to sample the field values $x \in \mathbb{R}^N$.  This
section explains why the obvious choices (random-walk MH, vanilla HMC) become
increasingly expensive as $N$ grows, and introduces two \emph{dimension-robust}
alternatives that are designed from the start to work on function space.

\subsection{The Curse of Discretisation}
\label{sec:curse-disc}

Fix a prior $x \sim \mathcal{N}(0, Q^{-1})$ with precision $Q \in \mathbb{R}^{N\times N}$.
As the mesh is refined, $N\to\infty$, and the field $x$ converges to a genuine
Gaussian process.  Standard samplers degrade:

\begin{itemize}
  \item \textbf{Random-walk MH} with isotropic proposal variance $h^2 I$:
        the optimal acceptance rate requires $h = O(N^{-1/2})$, so the chain
        makes $O(N)$ steps to move $O(1)$ in function space — cost $O(N)$ per
        independent sample.
  \item \textbf{Vanilla HMC} (mass matrix $M = I$): the leapfrog step size must
        satisfy $\varepsilon = O(N^{-1/4})$ to keep $|\Delta H|$ bounded, so
        $O(N^{1/4})$ steps per proposal and $O(N^{1/4})$ gradient evaluations,
        giving $O(N^{5/4})$ cost per independent sample.
\end{itemize}

The root cause is the same in both cases: the proposal ignores the prior geometry,
so the acceptance probability collapses in high dimensions.  The fix is to build
the prior covariance $Q^{-1}$ directly into the proposal.

\subsection{Preconditioned Crank--Nicolson (pCN)}
\label{sec:pcn}

\paragraph{Proposal.}
For a target $\pi(x) \propto L(x)\,\mathcal{N}(x;\,0,Q^{-1})$, define:
\begin{equation}
  x' = \sqrt{1-\beta^2}\,x + \beta\,\xi, \qquad \xi \sim \mathcal{N}(0, Q^{-1}),
  \label{eq:pcn-proposal}
\end{equation}
where $\beta \in (0,1]$ is a step-size parameter.

\paragraph{Why it works.}
The Crank--Nicolson interpolation \eqref{eq:pcn-proposal} exactly preserves
$\mathcal{N}(0, Q^{-1})$: if $x$ is distributed as the prior, so is $x'$.
Consequently, the prior log-density cancels in the Metropolis--Hastings ratio and
only the likelihood ratio survives:
\begin{equation}
  \alpha(x, x') = \min\!\left(1,\, \frac{L(x')}{L(x)}\right).
  \label{eq:pcn-alpha}
\end{equation}
The acceptance rate therefore depends only on $L$, not on $N$: choosing $\beta$
once for a coarse mesh yields the same acceptance on a fine mesh.  This
\emph{dimension-robustness} is the defining property of pCN
\citep{cotter2013}.

\paragraph{Implementation.}
Sampling $\xi \sim \mathcal{N}(0, Q^{-1})$ amounts to solving the sparse linear
system $Q\,\xi = z$ for $z \sim \mathcal{N}(0,I)$, which can be done with
conjugate gradient (CG) in $O(N)$ time for a graph-Laplacian precision (see
Section~\ref{sec:cg}).

\subsection{Hilbert-Space HMC}
\label{sec:hilbert-hmc}

pCN is robust but uses only first-order (likelihood evaluation) information.
Hilbert-space HMC incorporates gradient information while retaining
dimension-robustness \citep{beskos2011}.

\paragraph{Velocity parameterisation.}
Standard HMC introduces a momentum $p \sim \mathcal{N}(0, M)$ and
Hamiltonians $H = U(x) + \tfrac{1}{2}p^\top M^{-1} p$.  Setting $M = Q$
and changing variables to $v = Q^{-1}p$ (so $v \sim \mathcal{N}(0, Q^{-1})$):
\begin{align}
  H(x,v) &= U(x) + K(v), \\
  U(x)   &= -\log L(x) + \tfrac{1}{2} x^\top Q x, \label{eq:hmc-U}\\
  K(v)   &= \tfrac{1}{2} v^\top Q v. \label{eq:hmc-K}
\end{align}
Hamilton's equations become $\dot{x} = v$ and $Q\dot{v} = -\nabla U(x)$.

\paragraph{Leapfrog integrator.}
With step size $\varepsilon$ and $L$ leapfrog steps:
\begin{align}
  v &\leftarrow v - \tfrac{\varepsilon}{2}\, Q^{-1}\nabla U(x) \tag{half kick}\\
  \text{for } & i = 1,\ldots, L-1: \notag\\
  & x \leftarrow x + \varepsilon\, v \tag{drift}\\
  & v \leftarrow v - \varepsilon\, Q^{-1}\nabla U(x) \tag{full kick}\\
  x &\leftarrow x + \varepsilon\, v \tag{final drift}\\
  v &\leftarrow v - \tfrac{\varepsilon}{2}\, Q^{-1}\nabla U(x). \tag{final half kick}
\end{align}
Each ``kick'' requires one solve $Q^{-1}b$; each drift requires only a
vector addition.  The kinetic energy $K(v) = \tfrac{1}{2}v^\top Qv$ needs only
a matrix–vector product $Qv$, not a solve.

\paragraph{Why the velocity form is preferable.}
Three advantages over the standard $(x, p)$ form with $M = Q$:
\begin{enumerate}
  \item $K(v) = \tfrac{1}{2}v^\top Qv$ — computed with one sparse mat-vec, no
        matrix inverse.
  \item Velocity refresh $v \sim \mathcal{N}(0, Q^{-1})$ uses \emph{exactly} the same
        CG solve as prior sampling — no Cholesky, no dense matrix stored.
  \item All $O(N^2)$--$O(N^3)$ dense-algebra is replaced by $O(N \times
        \text{CG iters})$ sparse work.
\end{enumerate}

\paragraph{Dimension-robustness.}
With $M = Q$, the leapfrog step size can be chosen independently of $N$ (the
eigenmodes of $Q$ all enter the dynamics on the same footing).  In practice
one observes $|\Delta H| = O(1)$ and acceptance $\approx 65$--$80\,\%$ regardless
of mesh size, provided $\varepsilon$ is tuned on a coarse mesh \citep{beskos2011,
hairer2014}.

\paragraph{CG for $Q^{-1}b$.}
\label{sec:cg}
For the Besag CAR prior $Q = \kappa^2 I + L_{\rm graph}$, where $L_{\rm graph}$
is the graph Laplacian of the mesh connectivity, the condition number satisfies
\begin{equation}
  \kappa(Q) = \frac{\kappa^2 + \lambda_{\max}(L_{\rm graph})}{\kappa^2}
            \approx \frac{\kappa^2 + 8}{\kappa^2}
\end{equation}
for a 4-connected grid (eigenvalues of $L_{\rm graph}$ lie in $[0,8]$).  This is
\emph{independent of $N$}, so CG converges in $O(1)$ iterations and each leapfrog
step costs $O(N)$.  For the FEM precision matrix derived from the SPDE operator
(Section~\ref{sec:precision}), the same argument holds with $\kappa^2$
replacing 8 as the dominant eigenvalue bound.

\subsection{Combining Hilbert-Space HMC with RJMCMC}
\label{sec:rjmcmc-hmc}

The overall algorithm separates into two nested move types:

\begin{description}
  \item[\textbf{Within-model move}] (fixed $N_m$): one Hilbert-space HMC step in
    $(x, v)$ space.  Gradient information from the likelihood accelerates mixing;
    the $Q$-preconditioned leapfrog keeps $\varepsilon$ mesh-size-independent.
  \item[\textbf{Transdimensional move}] (birth or death): one RJMCMC step as in
    Section~\ref{sec:rjmcmc}.  On acceptance, the field $x$ and precision $Q$
    change dimension.  The velocity $v$ is refreshed by drawing
    $v_{\rm new} \sim \mathcal{N}(0, Q_{\rm new}^{-1})$
    (one CG solve in the new space) before the next HMC step.
\end{description}

\begin{algorithm}[H]
\caption{Joint RJMCMC + Hilbert-Space HMC}
\label{alg:rj-hmc}
\begin{algorithmic}[1]
  \State \textbf{Initialise:} mesh $\mathcal{T}_0$, $x^{(0)} \sim \mathcal{N}(0, Q_0^{-1})$,
         $v^{(0)} \sim \mathcal{N}(0, Q_0^{-1})$
  \For{$t = 1, 2, \ldots$}
    \State Draw $u \sim \mathrm{Uniform}(0,1)$
    \If{$u < p_{\rm trans}$}  \Comment{transdimensional move}
      \State Propose birth/death $(\mathcal{T}, x) \to (\mathcal{T}', x')$ \hfill (Section~\ref{sec:rjmcmc})
      \State Accept with probability $\min(1,\, \mathcal{R}_L \cdot \mathcal{R}_m \cdot \mathcal{R}_J)$
      \If{accepted}
        \State $v \leftarrow Q'^{-1} z$, \;$z \sim \mathcal{N}(0, I)$
        \hfill (velocity refresh via CG)
      \EndIf
    \Else \Comment{within-model move}
      \State $v \leftarrow Q^{-1}z$, \;$z \sim \mathcal{N}(0, I)$
            \hfill (velocity refresh)
      \State Run $L$-step leapfrog to propose $(x', v')$
      \State Accept with probability $\min(1,\, e^{-\Delta H})$
    \EndIf
  \EndFor
\end{algorithmic}
\end{algorithm}

\noindent
The mixing-time scales of the two move types are naturally separated: HMC
explores the field posterior quickly (correlated draws in $O(1)$ steps) while
RJMCMC explores mesh complexity slowly (each birth/death requires a Jacobian
factor).  The fraction $p_{\rm trans}$ controls the trade-off; a common heuristic
is $p_{\rm trans} \approx 0.1$.

\subsection{Annotated Reading List}
\label{sec:reading-list}

\begin{description}
  \item[\citet{stuart2010}] The foundational reference for Bayesian inversion on
    function space.  Section 4 defines Gaussian priors on Hilbert spaces and makes
    precise what it means for a posterior to be well-posed in infinite dimensions.
    Read this first to understand \emph{why} dimension-robustness is the right
    criterion.

  \item[\citet{cotter2013}] Derives pCN and its acceptance formula \eqref{eq:pcn-alpha},
    proves dimension-robustness, and compares it to a function-space version of
    the Langevin algorithm (MALA).  Includes numerical experiments on an elliptic
    inverse problem showing acceptance rates stable as $N \to \infty$.

  \item[\citet{beskos2011}] Introduces the Hilbert-space HMC (called ``Hybrid Monte
    Carlo on Hilbert spaces'').  Proves that with the prior as mass matrix the
    step size can be held fixed as $N \to \infty$ and that the acceptance
    probability remains bounded away from zero.  The velocity parameterisation
    used here follows their exposition.

  \item[\citet{hairer2014}] Provides spectral-gap bounds for pCN and related
    infinite-dimensional Metropolis algorithms, giving a rigorous convergence rate
    in terms of the likelihood Lipschitz constant.  Useful background if you want
    to understand \emph{how fast} the chain mixes in theory.

  \item[\citet{neal2011}] The standard handbook chapter on HMC; covers the
    leapfrog integrator, step-size tuning, and the dual-averaging adaptation used
    in NUTS.  Read before \citet{beskos2011} to understand the finite-dimensional
    version.

  \item[\citet{green1995}] Original RJMCMC paper.  Introduces the general
    reversible-jump framework and the Jacobian term $\mathcal{R}_J$; the birth/death
    moves of Section~\ref{sec:rjmcmc} are a special case.
\end{description}

% ============================================================
\section{Implementation Summary}
% ============================================================

The pipeline is split across two Python modules and one Jupyter notebook:

\begin{description}
  \item[\texttt{triangulation.py}] Adaptive kd-tree mesh construction
    (\texttt{AdaptiveKDMesh}) and vectorised FEM assembly
    (\texttt{assemble\_fem\_matrices}, \texttt{precision\_matrix}).
    Supports isotropic and anisotropic diffusion tensors $H$.

  \item[\texttt{inference.py}] Log-posterior construction
    (\texttt{make\_log\_posterior}, \texttt{make\_log\_posterior\_with\_hyperparams}),
    NUTS runner (\texttt{run\_nuts}), and posterior prediction
    (\texttt{predict\_intensity}).

  \item[\texttt{spde\_triangulation.ipynb}] Demonstration on synthetic data from
    a Gaussian mixture Poisson process.  Compares fixed-$\kappa$ and
    joint-hyperparameter models; visualises posterior median intensity,
    point-wise uncertainty, and parameter traces.
\end{description}

% ============================================================
\section{Discussion and Extensions}
% ============================================================

\subsection{Extensions Already Scaffolded}

The \texttt{assemble\_fem\_matrices} function accepts a callable $H(s)$
returning a $(T \times d \times d)$ array of per-simplex diffusion tensors,
enabling \emph{non-stationary} anisotropy.  Section~\ref{sec:anisotropic}
covers the stationary anisotropic case in detail; the key remaining obstacle
for both cases is the $O(m^3)$ log-determinant (Section~\ref{sec:logdet-aniso}).
Likewise, $\kappa(s)$ can be a callable returning per-simplex values; the
eigenvalue log-det trick then no longer applies.

\subsection{Planned: Non-Gaussian Fields (Level 2)}

The current LGCP assumes a Gaussian field, which may be unsuitable for intensity
functions with sharp discontinuities.  Planned extensions include:
\begin{itemize}
  \item \textbf{$\alpha$-stable SPDE}: replace Gaussian white noise $\mathcal{W}$
    with an $\alpha$-stable L\'evy sheet, yielding a Cauchy or heavy-tailed field;
  \item \textbf{Jump-diffusion SPDE}: add a compound Poisson term to the SPDE,
    allowing finite jumps over infinitesimal regions.
\end{itemize}

Both extensions require replacing the Gaussian prior log-density in
\eqref{eq:log-posterior} with a non-Gaussian counterpart and potentially
switching to a different sampler (e.g., elliptical slice sampling or particle
MCMC).

\subsection{Scalability}

The FEM assembly is fully vectorised (no Python loops over simplices) and runs
in seconds for meshes up to $\sim 10^4$ nodes.  For the isotropic model, the
dominant cost is the one-time eigendecomposition (Section~\ref{sec:logdet}),
$O(m^3)$, taking $\sim 1$\,s for $m \approx 500$; each subsequent NUTS step
costs $O(m)$.  For the anisotropic model, the $O(m^3)$ \texttt{slogdet} is paid
\emph{every leapfrog step} (Section~\ref{sec:logdet-aniso}), limiting the
practical mesh size to $m \lesssim 500$.  For larger isotropic meshes one would
switch to a sparse Cholesky factorisation (e.g., via \textsc{CHOLMOD}) and
accept $O(m)$ per-step cost without the precomputed eigenvalues.

\bibliographystyle{plainnat}
\bibliography{refs}

\end{document}
